Variance reduction for MCMC methods via martingale representations

### 2020-02-20 aap/Revision/vr-mcmc-revision-new.tex
update vr-mcmc-revision-new.tex
theorem 1 update (not final)

$\phi_k(\xi)$, $k\ge1$, are centered. As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of Shifted Legendre polynomials for "uniform part" and Hermite polynomials for "Gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the Shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\) We shall use the notations $\nset=\{1,2,\ldots\}$ and $\nset_0=\mathbb N\cup\{0\}$. We denote by $(\mathcal{G}_p)_{p\in \mathbb{Z}_+}$  the filtration generated by independent random variables $(\xi_p)_{p \in \mathbb{N}^*}$ with the convention $\mathcal{G}_0=\mathrm{triv}$. For $l \le p$ and $x \in \rset^d$, put for $x,y \in \rset^d$ functions $\Phi_k(x,y)=x-\gamma_k\mu(x)+y$. Then we can define

@@

with the function \(G_{l,p}:\) \(\mathbb{R}^{d\times(p-l+2)}\to \mathbb{R}^{d}\) is defined as

@@

For simplicity we will write $X^x_{p}$ and $G_{p}$ instead of $X^x_{0,p}$ and $G_{0,p}$. As a consequence, for a function $f\colon \rset^d \to \rset$ we have

@@

f\circ G_{p}(x,\sqrt{\gamma}\xi_{1},\ldots,\sqrt{\gamma}\xi_{p})

@@

 Let $f$ be a Borel function $\rset^{d}\to\mathbb R$ such that
$\mathsf{E}\left[\left |f(X_{p})\right|^{2}\right]<\infty$,   where $(X_{p})_{p\geq 0}$ is a Markov chain of the form \eqref{eq:chain_gen} for some sequence of functions $(\Phi_l)_{l\geq 1}.$
Then,
for $p>j$, the following representation holds in \(L^2(P)\)

@@

a_{p,l,k}(x)=\mathsf{E}\left[f(X^x_{l-1,p})\phi_k\left(\xi_{l}\right)\right], \quad p\geq l, \quad k\in \mathbb{N}

@@

\begin{proof}%[Proof of Theorem~\ref{prop:29032018a1}]
The expansion obviously holds for $p=1$ and $j=0$.
Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system in \(L^2(\mathbb{R}^d, P_{\xi})\), we have for all $x \in \rset^d$,
\eric{it is clear, but we must say in which sense this equality should be understood}
\[
f(X_{1})=\PE[f(X_{1})]+\sum_{k\geq1}a_{1,1,k}(x)\phi_{k}(\xi_{1})
\]
with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$,
provided $\PE[|f(X_{1})|^{2}]<\infty.$
\tcr{Recall that $\mathcal{G}_l=\sigma(\xi_{1},\ldots,\xi_{l}),$
$l=1,2,\ldots, $
and $\mathcal{G}_0=\mathrm{triv}$.}
\eric{do we need this, this is recalled ten lines above}
Suppose that (\ref{eq:mart_repr})
holds for $p=q$, all $j<q$, and all Borel-measurable functions
$f$ with $\mathsf{E}\left[|f(X_{q})|^2\right]<\infty$.
Let us prove it for $p=q+1$.
Given $f$ with $\mathsf{E}\left[|f(X_{p})|^2\right]<\infty$,
due to the orthonormality and completeness
of the system $\left(\phi_{k}\right)$, we get by
conditioning on $\mathcal{G}_{q}$,
\[
f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|{\mathcal{G}}_q\right]+\sum_{k\geq1}\alpha_{p,q+1,k}\phi_{k}(\xi_{q+1}),
\]
where
\begin{equation*}
\alpha_{p,q+1,k}
=\mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{q+1})\right|\mathcal G_q\right].
\end{equation*}
By the Markov property of $(X_{l})$,
we have
$\mathsf{E}[f(X_{p})|\mathcal{G}_q]
=\mathsf{E}[f(X_{p})|X_{q}]$.
Furthermore, a calculation involving
intermediate conditioning on $\mathcal G_{q+1}$
and the recurrence relation
$X_{q+1}=\Phi_{q+1}(X_{q},\xi_{q+1})$
verifies that
\[
\alpha_{p,q+1,k}=\mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{q+1})\right|X_{q}\right]
=a_{p,q+1,k}(X_{q})
\]
for suitably chosen
Borel-measurable functions $a_{p,q+1,k}$.
We thus arrive at
\begin{align}
\label{berm:sig_X}
f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|X_{q}\right]+\sum_{k\geq1}a_{p,q+1,k}(X_{q})\phi_{k}(\xi_{q+1}),
\end{align}
which is the required statement in the case $j=q$.
Now assume $j<q$.
The random variable
$\mathsf{E}\left[\left.f(X_{p})\right|X_{q}\right]$
is square integrable and has the form
$g(X_{q})$,
hence the induction hypothesis applies, and we get
\begin{equation}\label{eq:28082017a1}
\mathsf{E}\left[\left.f(X_{p})\right|X_{q}\right]=\mathsf{E}\left[\left.f(X_{p})\right|X_{j}\right]+\sum_{k\geq1}\sum_{l=j+1}^{q}a_{p,l,k}(X_{l-1})\phi_{k}(\xi_{l})
\end{equation}
with
\begin{align*}
a_{p,l,k}(X_{l-1}) &= \mathsf{E}\left[\left.\mathsf{E}\left[\left.f(X_{p})\right|\mathcal{G}_{q}\right]\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]
= \mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]\\
&=\mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{l})\right|X_{l-1}\right].
\end{align*}
Formulas \eqref{berm:sig_X}
and~\eqref{eq:28082017a1} conclude the proof.
\end{proof}

@@

\section{Numerical results}
\label{sec:num}
We will perform numerical experiments based on ULA with constant step size $\gamma$. Suppose that each function  \(Q_{r}(x)\) from \ref{eq:definition-bar-a} can be well approximated by polynomials, and we constructed a polynomial approximation  for each \(Q_r(x)\)  in the form:
\begin{equation*}
\widehat{Q}_{r}(x) = \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} x^{\mathbf{s}},\quad s=(s_1,\ldots,s_d)
\end{equation*}
with some coefficients \(\beta_{\mathbf{s}}\in \mathbb{R}.\)
Then using the identity
\[
\xi^j = j! \sum_{r = 0}^{\lfloor j/2 \rfloor} \frac{1}{2^r  r! \sqrt{(j-2r)!}} H_{j-2r}(\xi),\quad \xi \in \mathbb{R},
\]
we obtain closed-form expression for the estimates $\widehat a_{r,k}(x)$ of functions $\bar a_{r,k}(x)$ from \ref{eq:definition-bar-a}. Namely, for all $x \in \mathbb{R}^d$, 
\begin{eqnarray*}
\widehat a_{r,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(x) \widehat{Q}_{r}(x - \gamma \mu(x) + \sqrt{\gamma}\xi)}{\TrainSet \vee \mathcal{G}_N}
= \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{i,k_i,s_i}(x),
\end{eqnarray*}
where for all integers $i,k_i,s_i$ and $x \in \rset^d$,
\begin{align*}
P_{r,k,s}(x) &= \mathsf{E} \left[ H_{k}(\xi_l) (x - \gamma \mu(x) + \sqrt{\gamma} \xi)^{s} \right]
\end{align*}
is a one-dimensional polynomial (in \(x\)) of degree at most \(s\) with analytically known coefficients. In consequence, it is enough to estimate functions \(Q_{r}(x)\). It can be done using a modified least-squares criteria based on $T$ training trajectories \(\left(X_1^{(s)},\ldots,X_{N+n}^{(s)}\right), s = 1,\ldots,T\) with $N$ being the size of burn-in period:
\begin{eqnarray}
\label{eq:Qcircr}
\widehat{Q}_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
\end{eqnarray}
for \(1 \leq r \leq n-1,\) $\Psi$ being the class of polynomials $\Psi = \{\psi(x) | \psi(x) = \sum_{\|\mathbf{s}\|\leq m} \alpha_{\mathbf{s}} x^{\mathbf{s}} \}$ and $\widehat{Q}_0(x) = f(x)$ by definition. Due to Lemma~\ref{a_decay},
\sergey{we need to put some lemma about the truncation point here, I will do it}
it is enough to estimate $Q_{r}$ for $r < n_{\mathrm{trunc}}$ for some truncation level \(n_{\mathrm{trunc}}\) depending on \(d\) and \(\gamma.\) It allows us to use a smaller amount of training trajectories to approximate $Q_{r}(x)$.

Finally  we construct a truncated version of the estimator from \ref{}:
\[
\pi_{K,n,n_{\mathrm{trunc}}}^N(f) = \pi_n^N(f) -  \widehat M_{K,n, n_{\mathrm{trunc}}}^{N}(f),
\]
where
\begin{eqnarray*}
\widehat M_{K,n,n_{\mathrm{trunc}}}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{0 < \|\mathbf{k}\| < K}\sum_{l=N+1}^{p} \widehat a_{p-l,\mathbf{k}}(X_{l-1})\phi_k(\xi_{l})\mathbbm{1}\{ | p-l | < n_{\mathrm{trunc}}\right]
\end{eqnarray*}

\subsection{Gaussian mixtures}
 We consider a sample generated by ULA with $\pi$  given by the mixture of two Gaussian distributions  with equal weights:
\[
\pi (x) = \frac{1}{2(2\pi)^{d/2}} \left( \rme^{\frac{-\| x-a\|_2^2}{2}}  + \rme^{\frac{-\| x+a\|_2^2}{2}} \right), \quad x \in\mathbb R^d
\]
where $a \in \mathbb{R}^d$ is a given vector. The function $U(x)$ and its gradient are given by
\[
U(x) = \frac{1}{2} \|x - a\|_2^2 - \text{log}(1 + \rme^{-2x^\top a}), \,\,
\nabla U(x) = x-a +2a(1 + \rme^{2 x^\top a})^{-1},
\]
respectively.
In our experiments we considered dimensions $d = 2$ and $d = 8$ and take $a=( (2d)^{-1/2}, \dots, (2d)^{-1/2})^\top$.
In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i,\)
 we have used constant step sizes $\gamma_i = \gamma=0.2$ and sampled $n = 10^4$ samples for $d = 2$ and $n = 2 \times 10^3$ for $d = 8$.  We generated $\NtrainPath = 10$ independent "training" trajectories and solved the least squares  problems \eqref{eq:Qcircr} using the first order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section. The truncation level \(n_{\mathrm{trunc}}\) is chosen to be \(50.\) To test our variance reduction algorithm, we generated 100 independent trajectories.
 In Figure ~\ref{fig:1} we compare our approach to variance reduction methods of \cite{mira2013zero} and \cite{brosse2018diffusion}.

>>>>

$\phi_k(\xi)$, $k\ge1$, are centered. As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of Shifted Legendre polynomials for "uniform part" and Hermite polynomials for "Gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the Shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\) We shall use the notations $\nset=\{1,2,\ldots\}$ and $\nset_0=\mathbb N\cup\{0\}$. We denote by $(\mathcal{G}_p)_{p\in \mathbb{Z}_+}$  the filtration generated by independent random variables $(\xi_p)_{p \in \nset}$ with the convention $\mathcal{G}_0=\mathrm{triv}$. For $l \le p$ and $x \in \rset^d$, put for $x,y \in \rset^d$ functions $\Phi_k(x,y)=x-\gamma_k\mu(x)+y$. Then we can define
\label{eq:ula_new}

@@

with the function \(G_{l,p}:\) \(\mathbb{R}^{d\times(p-l+2)}\to \mathbb{R}^{d}\) defined as

@@

Note that $\left(X^x_{0,p}\right)_{p \in \nset_0}$ is a Markov chain of the form \eqref{eq:chain_gen}, starting at $X_0 = x$. For simplicity we write $X^x_{p}$ and $G_{p}$ instead of $X^x_{0,p}$ and $G_{0,p}$. As a consequence, for a function $f\colon \rset^d \to \rset$ we have

@@

f\circ G_{p}(x,\sqrt{\gamma_1}\xi_{1},\ldots,\sqrt{\gamma_p}\xi_{p})

@@

Let \((X^x_{p})_{p\geq 0}\) be a Markov chain of the form \eqref{eq:ula_new}  and $f$ be a Borel function $\rset^{d}\to\mathbb R$ such that
$\mathsf{E}\left[\left |f(X^x_{p})\right|^{2}\right]<\infty$. Then, for $p>j$, the following representation holds in \(L^2(P)\)

@@

a_{p,l,k}(y)=\mathsf{E}\left[f(X^y_{l-1,p})\phi_k\left(\xi_{l}\right)\right], \quad p\geq l, \quad k\in \mathbb{N}

@@

\begin{proof}%[Proof of Theorem~\ref{prop:29032018a1}]
The expansion obviously holds for $p=1$ and $j=0$.
Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system in \(L^2(\mathbb{R}^d, P_{\xi})\), we have for arbitary $x \in \rset^d$,
\eric{it is clear, but we must say in which sense this equality should be understood}
\[
f(X^x_{1})=\PE[f(X^x_{1})]+\sum_{k\geq1}a_{1,1,k}(x)\phi_{k}(\xi_{1})
\]
with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$,
provided $\PE[|f(X_{1})|^{2}]<\infty.$
\tcr{Recall that $\mathcal{G}_l=\sigma(\xi_{1},\ldots,\xi_{l}),$
$l=1,2,\ldots, $
and $\mathcal{G}_0=\mathrm{triv}$.}
\eric{do we need this, this is recalled ten lines above}
Suppose that (\ref{eq:mart_repr})
holds for $p=q$, all $j<q$, and all Borel-measurable functions
$f$ with $\mathsf{E}\left[|f(X_{q})|^2\right]<\infty$.
Let us prove it for $p=q+1$.
Given $f$ with $\mathsf{E}\left[|f(X_{p})|^2\right]<\infty$,
due to the orthonormality and completeness
of the system $\left(\phi_{k}\right)$, we get by
conditioning on $\mathcal{G}_{q}$,
\[
f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|{\mathcal{G}}_q\right]+\sum_{k\geq1}\alpha_{p,q+1,k}\phi_{k}(\xi_{q+1}),
\]
where
\begin{equation*}
\alpha_{p,q+1,k}
=\mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{q+1})\right|\mathcal G_q\right].
\end{equation*}
By the Markov property of $(X_{l})$,
we have
$\mathsf{E}[f(X_{p})|\mathcal{G}_q]
=\mathsf{E}[f(X_{p})|X_{q}]$.
Furthermore, a calculation involving
intermediate conditioning on $\mathcal G_{q+1}$
and the recurrence relation
$X_{q+1}=\Phi_{q+1}(X_{q},\xi_{q+1})$
verifies that
\[
\alpha_{p,q+1,k}=\mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{q+1})\right|X_{q}\right]
=a_{p,q+1,k}(X_{q})
\]
for suitably chosen
Borel-measurable functions $a_{p,q+1,k}$.
We thus arrive at
\begin{align}
\label{berm:sig_X}
f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|X_{q}\right]+\sum_{k\geq1}a_{p,q+1,k}(X_{q})\phi_{k}(\xi_{q+1}),
\end{align}
which is the required statement in the case $j=q$.
Now assume $j<q$.
The random variable
$\mathsf{E}\left[\left.f(X_{p})\right|X_{q}\right]$
is square integrable and has the form
$g(X_{q})$,
hence the induction hypothesis applies, and we get
\begin{equation}\label{eq:28082017a1}
\mathsf{E}\left[\left.f(X_{p})\right|X_{q}\right]=\mathsf{E}\left[\left.f(X_{p})\right|X_{j}\right]+\sum_{k\geq1}\sum_{l=j+1}^{q}a_{p,l,k}(X_{l-1})\phi_{k}(\xi_{l})
\end{equation}
with
\begin{align*}
a_{p,l,k}(X_{l-1}) &= \mathsf{E}\left[\left.\mathsf{E}\left[\left.f(X_{p})\right|\mathcal{G}_{q}\right]\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]
= \mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]\\
&=\mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{l})\right|X_{l-1}\right].
\end{align*}
Formulas \eqref{berm:sig_X}
and~\eqref{eq:28082017a1} conclude the proof.
\end{proof}

@@

\section{Numerical results}
\label{sec:num}
We will perform numerical experiments based on ULA with constant step size $\gamma$. Suppose that each function  \(Q_{r}(x)\) from \ref{eq:definition-bar-a} can be well approximated by polynomials, and we constructed a polynomial approximation  for each \(Q_r(x)\)  in the form:
\begin{equation*}
\widehat{Q}_{r}(x) = \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} x^{\mathbf{s}},\quad s=(s_1,\ldots,s_d)
\end{equation*}
with some coefficients \(\beta_{\mathbf{s}}\in \mathbb{R}.\)
Then using the identity
\[
\xi^j = j! \sum_{r = 0}^{\lfloor j/2 \rfloor} \frac{1}{2^r  r! \sqrt{(j-2r)!}} H_{j-2r}(\xi),\quad \xi \in \mathbb{R},
\]
we obtain closed-form expression for the estimates $\widehat a_{r,k}(x)$ of functions $\bar a_{r,k}(x)$ from \ref{eq:definition-bar-a}. Namely, for all $x \in \mathbb{R}^d$, 
\begin{eqnarray*}
\widehat a_{r,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(x) \widehat{Q}_{r}(x - \gamma \mu(x) + \sqrt{\gamma}\xi)}{\TrainSet \vee \mathcal{G}_N}
= \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{i,k_i,s_i}(x),
\end{eqnarray*}
where for all integers $i,k_i,s_i$ and $x \in \rset^d$,
\begin{align*}
P_{r,k,s}(x) &= \mathsf{E} \left[ H_{k}(\xi_l) (x - \gamma \mu(x) + \sqrt{\gamma} \xi)^{s} \right]
\end{align*}
is a one-dimensional polynomial (in \(x\)) of degree at most \(s\) with analytically known coefficients. In consequence, it is enough to estimate functions \(Q_{r}(x)\). It can be done using a modified least-squares criteria based on $T$ training trajectories \(\left(X_1^{(s)},\ldots,X_{N+n}^{(s)}\right), s = 1,\ldots,T\) with $N$ being the size of burn-in period:
\begin{eqnarray}
\label{eq:Qregr}
\widehat{Q}_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
\end{eqnarray}
for \(1 \leq r \leq n-1,\) $\Psi$ being the class of polynomials $\Psi = \{\psi(x) | \psi(x) = \sum_{\|\mathbf{s}\|\leq m} \alpha_{\mathbf{s}} x^{\mathbf{s}} \}$ and $\widehat{Q}_0(x) = f(x)$ by definition. Certainly one may use another functional class $\Psi$ in $\label{eq:Qregr}$, but using polynomials allows to compute closed-form representation of $\widehat a_{r,\mathbf{k}} (x)$.
\par
Due to Lemma~\ref{a_decay},
\sergey{we need to put some lemma about the truncation point here, I will do it}
it is enough to estimate $Q_{r}$ for $r < n_{\mathrm{trunc}}$ for some truncation level \(n_{\mathrm{trunc}}\) depending on \(d\) and \(\gamma.\) It allows us to use a smaller amount of training trajectories to approximate $Q_{r}(x)$.

Finally  we construct a truncated version of the estimator \ref{eq:29032018a3}:
\[
\pi_{K,n,n_{\mathrm{trunc}}}^N(f) = \pi_n^N(f) -  \widehat M_{K,n, n_{\mathrm{trunc}}}^{N}(f),
\]
where
\begin{eqnarray*}
\widehat M_{K,n,n_{\mathrm{trunc}}}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{0 < \|\mathbf{k}\| < K}\sum_{l=N+1}^{p} \widehat a_{p-l,\mathbf{k}}(X_{l-1})\phi_k(\xi_{l})\mathbbm{1}\{ | p-l | < n_{\mathrm{trunc}}\right]
\end{eqnarray*}

\subsection{Gaussian mixtures}
 We consider a sample generated by ULA with $\pi$  given by the mixture of two Gaussian distributions  with equal weights:
\[
\pi (x) = \frac{1}{2(2\pi)^{d/2}} \left( \rme^{\frac{-\| x-a\|_2^2}{2}}  + \rme^{\frac{-\| x+a\|_2^2}{2}} \right), \quad x \in\mathbb R^d
\]
where $a \in \mathbb{R}^d$ is a given vector. The function $U(x)$ and its gradient are given by
\[
U(x) = \frac{1}{2} \|x - a\|_2^2 - \text{log}(1 + \rme^{-2x^\top a}), \,\,
\nabla U(x) = x-a +2a(1 + \rme^{2 x^\top a})^{-1},
\]
respectively.
In our experiments we considered dimensions $d = 2$ and $d = 8$ and take $a=( (2d)^{-1/2}, \dots, (2d)^{-1/2})^\top$.
In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i,\)
 we have used constant step size $\gamma=0.2$ and sampled $n = 10^4$ samples for $d = 2$ and $n = 2 \times 10^3$ for $d = 8$.  We generated $\NtrainPath = 10$ independent "training" trajectories and solved the least squares  problems \eqref{eq:Qcircr} using the first order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section. The truncation level \(n_{\mathrm{trunc}}\) is chosen to be \(50.\) To test our variance reduction algorithm, we generated 100 independent trajectories.
 In Figure ~\ref{fig:1} we compare our approach to variance reduction methods of \cite{mira2013zero} and \cite{brosse2018diffusion}.
